




```{r}
library(dplyr)
library(magrittr)
```

```{r}

# loading the file 

escape <- read.csv('escapes.csv')
```


```{r}
# Dimensions o dataframes 

dim(escape)
```


the dataset contains of 357 records and 38 features . Now we take a look on summary

```{r}
# summary of the dataset
summary(escape)
```
I don't find summary very intresting now explore each feature 


now we look at the structure of data
```{r}
str(escape)
```
we can see most of the feature type is character

#Operator at the time of escape

we look at "Operator at the time of escape"

```{r}
levels((escape$Operator.at.Time.of.Escape))
```
it is showing null 

let's check it contains factor levels

```{r}
is.factor(escape$Operator.at.Time.of.Escape)
```

It means that columns is not a factor and does not have levels

```{r}
escape$Operator.at.Time.of.Escape <- as.factor(escape$Operator.at.Time.of.Escape)
levels(escape$Operator.at.Time.of.Escape)
```

it contain 79 different types of operator at the time of escapes

now let we check null value in this column 
```{r}
sum(is.na(escape$Operator.at.Time.of.Escape))


```

As it contains zero null values means this column is clean

#Escape Water Type

as this column in character data type,  first we need to convert it into factors then check levels and null values 

```{r}
escape$Escape.Water.Type <- as.factor(escape$Escape.Water.Type)
escape$Escape.Water.Type
levels(escape$Escape.Water.Type)
sum(is.na(escape$Escape.Water.Type))
```

as it contain only three type of values and it does not have any null value we can say that this column is also clean 

#Escape Start Date

as this column in character data type,  first we need to convert it into factors then check levels and null values

```{r}
escape$Escape.Start.Date <- as.POSIXct(escape$Escape.Start.Date, format="%d-%b-%y", tz="UTC")
escape$Escape.Start.Date

```
```{r}
sum(is.na(escape$Escape.Water.Type))
```


also in this column we don't find any null values and all the date convert to appropriate data type

#Escape Start Time 
```{r}
escape$Escape.Start.Time 
```


as we can see there are several values which does not state time 

```{r}
escape$Escape.Start.Time <- as.factor(escape$Escape.Start.Time)
escape$Escape.Start.Time
print('------')
levels(escape$Escape.Start.Time)
```
```{r}
#escape$Escape.Start.Time <- factor(c("04:00-0","10:00 am","10:00am","1000","11","11:00 am","1200","12pm","13","1430","17:00hr","1700","18","2pm","700","8.3","8.45","800","900"))

#levels(escape$Escape.Start.Time)[levels(escape$Escape.Start.Time) == "04:00-0"] <- "04:00"

escape <- escape %>% mutate(Escape.Start.Time = recode(Escape.Start.Time, "04:00-0" = "04:00", "10:00 am" = "10:00", "10:00am"= "10:00", "1000" = "10:00","11"="11:00", "11:00 am"="11:00","1200"="12:00","12pm"="12:00","13"="13:00","1430"="14:30","17:00hr"="17:00", "1700"="17:00","18"="18:00","2pm" = "14:00","700" = "07:00","8.3"="08:30","8.45"="08:45", "800"="08:00","900"="09:00"))
print("----")
levels(escape$Escape.Start.Time)
```

```{r}
escape <- escape %>% mutate(Escape.Start.Time = recode(Escape.Start.Time,"early h"= "NA","not kno" = "NA", "not pro" = "NA", "overnig"="NA", "pm"="NA","unknown"="NA", "n/a" = "NA"))

levels(escape$Escape.Start.Time)
```
we converted all the required values to appropriate values an


#Esca









